---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, instructions, eval = FALSE, include = FALSE}
# Build with
# load_all()
rmarkdown::render("README.Rmd")
```
<!-- which builds the .html that can be viewed locally (but isn't pushed to GitHub;
GitHub uses README.md to make the page you see on GitHub). See pacea if want to
save figures.
-->

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "60%"
)
library(tibble)
library(badger)
```

# sizeSpectra2

<!-- badges: start -->
`r badge_lifecycle("under development", "orange")`
[![R-CMD-check](https://github.com/andrew-edwards/sockeyePrize/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/andrew-edwards/sockeyePrize/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/andrew-edwards/sockeyePrize/branch/main/graph/badge.svg)](https://app.codecov.io/gh/andrew-edwards/sockeyePrize?branch=main)
![Visitors](https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fgithub.com%2Fandrew-edwards%2FsockeyePrize&label=VISITORS&countColor=%23263759&style=flat&labelStyle=lower)
<!-- badges: end -->

An R package for entering the SalmonPrize competition


## Installation

To install the latest version just:

```
install.packages("remotes")    # If you do not already have the "remotes" package

remotes::install_github("andrew-edwards/sockeyePrize")
```

If you get an error like
```
Error in utils::download.file(....)
```
then the connection may be timing out (this happens to me on our work network). Try

```
options(timeout = 1200)
```
and then try and install again. If you get a different error then post an Issue
or contact <a href="mailto:andrew.edwards@dfo-mpo.gc.ca">Andy</a> for help.
