---
title: "Wrangling the salmon data provided for 2026 SalmonPrize competition"
author: "Andrew Edwards and Carrie Holt"
output: pdf_document
fontsize: 12pt
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  cache = TRUE,
  cache_path = "data-2026-cache/",
  fig.path = "data-2026-figs-cache/",
  fig.width = 7,
  fig.height = 6
)
# comment = "#>",
knitr::dep_prev()
```

```{r, build, echo = FALSE, eval = FALSE}
# To build either run this line or click knit button in RStudio:
rmarkdown::render("data-2026.Rmd")
```

```{r, packages, echo = FALSE}
library(dplyr)
load_all()
```

TODO Didn't render properly but I saved data okay in package manually TODO

2026: Currently just have Columbia River values, so can't just rerun last year's
code. 

Looking at new `report-2026/columbia-analysis.Rmd` think we just need what was
in last year's `returns_all`. So just get that saved here, exiting before doing
the other wrangling that we needed last year for all stocks.

Okay, there are two files in
`data-for-2026-competition/Columbia_2026/1_SourceData/` (and more elsewhere):

- `ReturnsbyAge_COLUMBIA_2026.csv`

- `Spawners_COLUMBIA_2026.csv`.

Think we just want the first, hopefully columns etc. are same as last year.
Yes, it's the one with colummns like `AgeClass_0.2` etc.

Notes from that file:

1. Bonneville Lock & Dam is the return estimate to the lower river (after marine interceptions; but before in-river harvests)

1. Wenatchee and Okanagan stocks account for most of the total sockeye return to the Columbia

1. Wenatchee and Okanagan return estimates are based on estimated stock proportions
at Bonneville Lock & Dam

1. For other stocks this quantity might be labelled gross escapement or terminal run or return to river

1. For Wenatchee have replaced NA for some minor age classes with 0 in order to
allow for total recruits by brood year calculations


So last year there was just one River, this year three. Have to give 'All of
Columbia', 'Okanagan', and 'Wenatchee'.


Moving code up from below (this was `data-2025.Rmd` last year).

```{r, dataprep}
data_dir <- paste0(here::here(),
                   "/data-raw/data-for-2026-competition/Columbia_2026/1_SourceData/")

returns_all_raw <- read.csv(paste0(data_dir,
                                   "ReturnsbyAge_COLUMBIA_2026.csv"),
                            comment.char = "#") %>%
  tibble::as_tibble()

returns_all <- returns_all_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River),
                   factor)

returns_all

summary(returns_all)
```

This looks like that in `data-2025.Rmd`, except here is just for the one System,
but this one now has three Rivers.

```{r, savedata1}
returns_columbia_2026 <- returns_all
usethis::use_data(returns_columbia_2026)
```

## Effective spawners for Fraser River

Had to adapt from 2025 as they changed the data so need to use a different file.

Using the necessary code from 2025 (moving from below), just effective
spawners. Only available for FR, but filename looks the same (almost -- they
changed it from `...Form.csv` to `...Format.csv` to waste some of my time. Previously the 

THIS is not evaluated or needed, see next bit.
```{r, datapreparation77, eval = FALSE}
full_long_raw <- read.csv("data-for-2026-competition/DataPackage_Fraser_2026/FullDataSet_LongFormat.csv",
                          comment.char = "#") %>%
  tibble::as_tibble()

full_long_raw

effective_spawners_2026 <- full_long_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River,
                               Euro,
                               GR),
                   factor)

effective_spawners_2026

summary(effective_spawners_2026)
```
Not doing: Just FR. So save in package. Looks like they changed some column names also, but
deal with that in analysis.

```{r, datapreparation78}
full_long_raw <- read.csv("data-for-2026-competition/DataPackage_Fraser_2026/ALT_Spawners_EffectiveFemales_FRASER_2026.csv",
                          comment.char = "#") %>%
  tibble::as_tibble()

full_long_raw

effective_spawners_2026 <- full_long_raw %>%
  dplyr::rename(Late_Stuart = Late.Stuart)

effective_spawners_2026

summary(effective_spawners_2026)
```



Save data in package, doing manually (not evaluated) to not keep overwriting.
```{r, savedata3, eval = FALSE}
usethis::use_data(effective_spawners_2026, overwrite = TRUE)
```


```{r, stop, cache = FALSE}
knitr::knit_exit()
```



Wrangling the data. From instructions Word file:


To assist Salmon Prize teams in rapidly developing their estimates, we have pre-packaged brood and return tables for the fourteen runs across three systems. This includes:

1.	All 14 runs across the three systems
a.	Brood table
b.	Return table
c.	First year at sea table

2.	Bristol Bay System — covering eight runs ()
a.	Brood table
b.	Return Table

3.	Fraser River — covering five runs ()
a.	Brood table
b.	Return table

4.	Columbia River — one run covering the entire system
a.	Brood table
b.	Return table

These tables have been prepared by Alejandro Yanez. If anyone has any questions please reach out to Alejandro by email at a.yanez@oceans.ubc.ca.

--

So it seems that we can just get those in part 1. Will wrangle them a little and
make into data objects in our package.


# Data loading


```{r, datapreparation}
data_dir <- paste0(here::here(),
                   "/data-raw/data-for-2025-competition/Brood_Return_First_Year_At_Sea_Tables/")

brood_all_raw <- read.csv(paste0(data_dir,
                                 "Combined_Brood_Bristol_Columbia_Fraser.csv")) %>%
  tibble::as_tibble()

brood_all_raw

names(brood_all_raw)

brood_all <- brood_all_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River),
                   factor) %>%
  dplyr::select(-"X")

brood_all

summary(brood_all)
```


Now do same for returns and then first year at sea:
```{r, datapreparation2}
returns_all_raw <- read.csv(paste0(data_dir,
                                 "Combined_Return_Bristol_Columbia_Fraser.csv")) %>%
  tibble::as_tibble()

returns_all_raw

names(returns_all_raw)

returns_all <- returns_all_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River),
                   factor) %>%
  dplyr::select(-"X")

returns_all

summary(returns_all)
```

```{r, datapreparation3}
at_sea_all_raw <- read.csv(paste0(data_dir,
                                 "Combined_First_Year_At_Sea.csv")) %>%
  tibble::as_tibble()

at_sea_all_raw

names(at_sea_all_raw)

at_sea_all <- at_sea_all_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River),
                   factor) %>%
  dplyr::select(-"X")

at_sea_all

summary(at_sea_all)
```

## Updated data that now includes spawners

Just want spawners, looks like easy to get by brood year. Definition is:

`Total_Spawners_BroodYear: Number of spawners that produced this BroodYear's cohort`

There's also

`Total_EFS_BroodYear: Number of effective (after pre-spawn mortality) female
spawners that produced this year's cohort`

but that's only available by return year, so requires more fiddling. BUT really
should use, though it's only a Fraser River thing (thanks Catarina).

So need to do both - total spawners for all systems, and also (or instead) EFS
for FR.

### First total spawners, available for all Systems

First do the long form one which is by brood year, which we need for all
non-Fraser River (though can include it for FR).

```{r, datapreparation4}
data_dir_updated <- paste0(here::here(),
                           "/data-raw/data-for-2025-competition-updated/_Data_Updated_June26/Additional_Data_Spawners/Additional_Data_Spawners/")

spawners_brood_year_all_raw <- read.csv(paste0(data_dir_updated,
                                               "GENERATED_SRDataSet_LongForm.csv"),
                                        comment.char = "#") %>%
  tibble::as_tibble()

spawners_brood_year_all_raw

spawners_brood_year_all <- spawners_brood_year_all_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River),
                   factor)
spawners_brood_year_all

summary(spawners_brood_year_all)
```

So that will include recruits also, but we are using them from elsewhere since
dividing up by ages. Keep BroodYear here in name, to then do wrangling in the
data analysis.

### Now the more fiddly effective female spawners, which is only available for FR and is already saved differently

```{r, datapreparation5}
full_long_raw <- read.csv(paste0(data_dir_updated,
                                              "GENERATED_FullDataSet_LongForm.csv"),
                                       comment.char = "#") %>%
  tibble::as_tibble()

full_long_raw

effective_spawners_all <- full_long_raw %>%
  dplyr::mutate_at(dplyr::vars(System,
                               River,
                               Euro,
                               GR),
                   factor)

effective_spawners_all

summary(effective_spawners_all)
```

Check that only FR has `Total_EFS_BroodYear`:

```{r, datapreparation6}
summary(filter(effective_spawners_all, System != "Fraser River") %>%
        select(Total_EFS_BroodYear))
```
Indeed. So just save values for FR. Save everything and then extract what we
need in the data analysis.

```{r, datapreparation7}
effective_spawners <- filter(effective_spawners_all,
                             System == "Fraser River")

effective_spawners
summary(effective_spawners)
```

Save data in package, doing manually (not evaluated) to not keep overwriting.
```{r, savedata, eval = FALSE}
usethis::use_data(brood_all)
usethis::use_data(returns_all)
usethis::use_data(at_sea_all)
usethis::use_data(spawners_brood_year_all)
usethis::use_data(effective_spawners, overwrite = TRUE)
```
