---
title: "Wrangling North Pacific Salmon data - 2026 update"
author: "Andrew Edwards and Carrie Holt"
output: pdf_document
fontsize: 12pt
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  cache = FALSE,
  # cache_path = "data-2025-cache/",
  # fig.path = "data-2025-figs-cache/",
  fig.width = 7,
  fig.height = 6
)
  # comment = "#>",
```

```{r, build, echo = FALSE, eval = FALSE}
# To build either run this line or click knit button in RStudio:
rmarkdown::render("north-pacific-salmon-2026.Rmd")
```

```{r, packages, echo = FALSE}
library(dplyr)
load_all()
```

Updating for 2025, and saving data object as `north_pacific_salmon` instead of
`np_salmon`. 

Updated from `north-pacific-salmon.Rmd`.

From Brendan (30/4/25 email to Carrie):

"The attached are the most up to date data and best citation would be: Connors, B., Ruggerone, G.T. and Irvine, J.R., 2025. Adapting management of Pacific salmon to a warming and more crowded ocean. ICES Journal of Marine Science, 82(1), p.fsae135."

From the spreadsheet:

"data through 2015 taken from Ruggerone, G.T. and Irvine, J.R., 2018. Numbers
and biomass of natural and hatchery origin pink salmon, chum salmon, and sockeye
salmon in the North Pacific Ocean, 1925-2015. Marine and Coastal Fisheries,
10(2), pp.152-168. "

2016-2023 updated using same methods.

2024: crude estimate of pinks; 5 year average for chum and sockeye (presumably
would want to update for a formal paper, this is fine for our purposes).

From Brendan (10/10/26 email), summarising:

No 2025 update to time series yet, but attached is the most up to date dataset
through 2024 which may have some slight tweaks compared to previous file I
shared for values post 2015, including much more confident estimates in 2024. 
Can pass along 2025 estimates later this summer once commercial catch by nation is published by NPAFC.

# Data loading


```{r, datapreparation}
data_dir <- paste0(here::here(),
                   "/data-raw/")

total_raw <- read.csv(paste0(data_dir,
                             "total_np_salmon-2026-06-10.csv")) %>%
  tibble::as_tibble()

north_pacific_salmon <- mutate(total_raw,
                    return_year_of_np_salmon = Return.Year,
                    pink = Pink,
                    chum = Chum,
                    sockeye = Sockeye,
                    total_ocean_salmon = Total) %>%
  select(return_year_of_np_salmon,
         pink,
         chum,
         sockeye,
         total_ocean_salmon)

north_pacific_salmon

tail(north_pacific_salmon)

summary(north_pacific_salmon)
```

Save data in package:
```{r, savedata}
usethis::use_data(north_pacific_salmon,
                  overwrite = TRUE)
```
