---
title: "Creating Fraser River covariates to go into some pbsEDM analyses"
author: "Andrew Edwards and Carrie Holt"
output: pdf_document
fontsize: 12pt
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  cache = TRUE,
  cache_path = "fraser-covariates-2026-cache/",
  fig.path = "fraser-covariates-2026-figs-cache/",
  fig.width = 7,
  fig.height = 6
)

knitr::dep_prev()
```

```{r, build, echo = FALSE, eval = FALSE}
# To build either run this line or click knit button in RStudio:
rmarkdown::render("fraser-covariates-2026.Rmd")
```

```{r, packages, echo = FALSE, cache = FALSE}
load_all()
# Other users need to uncomment these two lines. 
# remotes::install_github("andrew-edwards/sockeyePrize")
# library(sockeyePrize)
library(dplyr)
library(pbsEDM)   # remotes::install_github("pbs-assess/pbsEDM")
library(pacea)    # remotes::install_github("pbs-assess/pbsEDM")
                  #   make sure to get latest update
library(gridExtra)
library(ggplot2)

# From Chris's hake
f <- function(x, digits = 0){

  if(is.null(x)){
    return(x)
  }

  format(round(x,
               digits),
         big.mark = ",",
         nsmall = digits)
}

```

Construct just the environmental covariates in 2026, adapting code that we used
in the 2025 salmon prize competition (code for 2025 was
`report/analysis-2025-3.Rmd`, just adapting the covariate part of that). 

These all come from pacea, except `north_pacific_salmon` which was obtained from
Brendan Rogers and saved into the current package.

These covariates (including things like the months to use for some of them) were
selected by Carrie and Andy in 2025 for Fraser River Chilko (we may have some justifications written up
elsewhere). 

Other can adapt this as they see fit, it should provide a good template.

Then see EDM analysis in `report-2026/`.

# Fraser River EDM - what we would put into pbsEDM.

Try to predict:

- age4 returns each year, say 2021 as an example. <!-- From `one_returns_age45`
  in original analysis-->

- age5 returns each year (2021), an additional covariate would be age4 returns.

Do each seperately and then give the sum as the total prediction.

So, age4 returns in 2021:

 2017: spawned in freshwater as age0, though think it's the winter, so could be
 early 2018

 2018: stay in freshwater as age1

 2019: outmigration to ocean as age2

 2020: at-sea as age3

 2021: return as age4

We assumed mostly 1.X, meaning they do 1 year in freshwater (think we looked at
the 0.X and 2.X and there were smaller). But just to note that the above life
cycle isn't true for all fish.

Environmental covariates would be (D = Done, N = not needed):

- PDO in the winter (Nov 2018 to Mar 2019) preceding outmigration for age4.

- PDO in the winter (average of Nov 2017 to Mar 2018 and Nov 2018 to Mar 2019)
  preceding outmigration for age5 for Chilko.

- PDO in the winter (Nov 2017 to Mar 2018) preceding outmigration for other
  age5's.

N Temperature from buoys: "West Sea Otter" (Apr-Jul average), "East Dellwood
  Knolls" (Apr-Jul average), "Halibut Bank" (Apr-Jun average). Year of
  outmigration, after the PDO Mar calculation, so 2019 for 2021 return year.
  Had thought that because hese did not work great (think maybe some missing data?), try the other buoys in SoG, now that I
  have the code done to look at fairly easily. But now doing OISST. 

- OISST - could either use Apr-Jul average for an area, or just pick the pixel
  closest to Halibut Bank buoy, which I think was our first choice.

- OISST - talking to Kaitlyn and Jessie, suggest looking at SST for the winter,
  as a proxy for other parts of the ecosystem (e.g. food availability for when
  fish returning). Possibly Dec-Feb, or rather do Nov-Mar, which we will do.

- peak (max daily value) and average daily Fraser River Discharge from Apr-Jun
  (same as buoy temperatures). Caveat now from SOPO meeting is that the Hope
  station is not so indicative of what's happening further down the river (I think).

- total pink salmon abundance in N Pacific. Second or third years of marine
  life of the sockeye (as competition). Year after the temperature data (second
  year), 2020.  `np_salmon` has `return_year_of_np_salmon`
  Carrie contacting Hannah Hunter -- think that informed Carrie's notes for the
  other two river systems below (even if it didn't, that's what we'll
  use). Brendan has sent us 2024, now in this package. See section below. Have
  asked Brendan and Dylan for 2025.

- total salmon abundance in N Pacific. Same as pink. Andy: though won't the
  non-pinks be out there for longer so we should look at multiple years? And
  don't think the years match up, see section below.

- copepods the year of outmigration (same as SST). Total biomass anomaly.

- bifurcation index, year of outmigration (same as SST). Not yet available from
  Mike Malick for 2025 though.


```{r, initialise}
start_year <- 1940           # Had taken from Chilko one_returns_age45, but just
                             #  pick a slightly earlier year here
end_year <- 2026              # Did have this as the year before the current
# year, but make it current year as will have things like winter PDO and winter
# SST for the current year. Work out the years to analyse in the analysis code.```
```

Had `Analysis for Fraser River Chilko` etc. (see `analysis-2025-3.Rmd` but removing here to just focus
on non-stock-specific covariates. 

### PDO in winter.

PDO in the winter (Nov 2018 to Mar 2019) preceding outmigration for age4. So in
EDM will expect a lag of 2 years.

Bit fiddly to group Nov to Mar, doing a quick kludge here but
should generalise in a function and add some checks. TODO at some point and put
into pacea. The year here corresponds to the year of the January in the middle
of the Nov-Mar average.

```{r pdo}
winter_months <- c(1, 2, 3, 11, 12)
pdo_winter_months <- filter(pdo, month %in% winter_months)
pdo_winter_months <- pdo_winter_months[-(1:3), ]    # need to generalise
pdo_winter_months <- cbind(pdo_winter_months,
                           winter_group = rep(1:(length(unique(pdo_winter_months$year)) - 1),
                                              each = length(winter_months))) %>%
  as_tibble()    # fails if not jan-mar in final year; need to ;
# generalise TODO
pdo_winter_months

pdo_winter <- summarise(group_by(pdo_winter_months,
                                 winter_group),
                        year = max(year),
                        pdo_winter_mean = mean(anomaly)) %>%
  select(-c("winter_group"))

pdo_winter %>% tail()
```

So we already have the `pdo_winter_mean` for 2026, but EDM should not need it as
we expect the influence to be from two years before (for 2025 age4 returns)  during outmigration.

### Temperature in spring -- from OISST

Just try pixel closest to Halibut Bank buoy, as can adapt code from Travis's vignette
  (and maybe functionalise it later).

From Travis's oisst.Rmd:

```{r}
lat <- buoy_metadata$latitude[which(buoy_metadata$name == "Halibut Bank")]
lon <- buoy_metadata$longitude[which(buoy_metadata$name == "Halibut Bank")]

# create a dataframe and convert to sf object
coords_halibut_bank <- data.frame(x = lon, y = lat)
sf_halibut_bank <- sf::st_as_sf(coords_halibut_bank,
                                coords = c("x", "y"),
                                crs = "EPSG: 4326")
sf_halibut_bank

# distance from buoy to each coordinate
dist <- sf::st_distance(oisst_month, sf_halibut_bank)

# subset data for points that are closest to buoy
subset_dat <- oisst_month[which(dist == min(dist)),]

subset_dat    # geometry is just the same point repeated I think
```

Now we can plot the time series for the data we've extracted

```{r, oisstallyears}
subset_dat %>%
  mutate(year = as.factor(year)) %>%  # set year to a factor so each line is plotted separately
  ggplot() +
  geom_line(aes(x = month, y = sst, col = year)) +
  scale_y_continuous(name = attributes(oisst_month)$units)
```

To see where we are this year, let's plot just this year's data, with the
historical distribution (not evaluating, year 2023 needs updating to
`max(subset_dat$year)` but didn't work, ignoring).

```{r, oisstoneyear, eval = FALSE, echo = FALSE}
# calculate summary statistics (median, sd, 0.05 and 0.95 probabilities)
summary_dat <- subset_dat %>%
  group_by(month) %>%
  summarise(median_val = median(sst, na.rm = TRUE),
            sd_val = sd(sst, na.rm = TRUE),
            q05 = quantile(sst, probs = 0.05, na.rm = TRUE),
            q95 = quantile(sst, probs = 0.95, na.rm = TRUE))

subset_dat %>%
  filter(year == 2023) %>%
  mutate(year = as.factor(year)) %>%
  ggplot() +
  geom_ribbon(data = summary_dat, aes(x = month, ymin = q05, ymax = q95), fill = "grey") +
  geom_line(aes(x = month, y = sst), col = "red") +
  geom_line(data = summary_dat, aes(x = month, y = median_val), col = "black", linewidth = 1) +
  scale_y_continuous(name = attributes(oisst_month)$units)
```

Now to extract what we need, the annual average over Apr-Jul for each
year. Adapting from PDO code above (be nice to have a function!):

```{r, oisstavge}
spring_months <- 4:7
halibut_bank_spring_months <- dplyr::filter(subset_dat, month %in% spring_months)

# Check the number of samples in each month:
min(halibut_bank_spring_months$sst_n)     # 30, great.

# Check first year has all four months (i.e. starts in April);
halibut_bank_spring_months    # Yes it does

halibut_bank_spring <- summarise(group_by(halibut_bank_spring_months,
                                          year),
                       year = max(year),
                       sst_spring = mean(sst)) %>%
  sf::st_drop_geometry()

halibut_bank_spring

halibut_bank_spring %>% tail()

plot(halibut_bank_spring$year,
     halibut_bank_spring$sst_spring,
     type = "o",
     xlab = "Spring SST",
     ylab = "Year")
```

### Temperature in winter -- from OISST

Using Nov-Mar as suggested, which is also the same as for PDO in winter. Convention is labelling with the year of the January, as for PDO
(and from discussion with oceanographers).

```{r wintertemp}
# Adapting from winter PDO and spring SST above
expect_equal(winter_months,
             c(1, 2, 3, 11, 12))  # in case it gets changed above

halibut_bank_winter_months <- dplyr::filter(subset_dat, month %in% winter_months) %>%
  sf::st_drop_geometry()

# Check the number of samples in each month:
min(halibut_bank_winter_months$sst_n)     # 27, good enough

# Just missing the one day
filter(halibut_bank_winter_months,
       sst_n < 28)

# Check all years (Nov-Dec) will be complete (yes):
halibut_bank_winter_months

halibut_bank_winter_months %>% tail()

halibut_sst_winter_grouped <- cbind(halibut_bank_winter_months,
                                    winter_group = rep(1:(length(unique(halibut_bank_winter_months$year)) - 1),
                                              each = length(winter_months))) %>%
  as_tibble()    # fails if not jan-mar in final year; need to ;
# generalise TODO
halibut_sst_winter_grouped
halibut_sst_winter_grouped %>% tail()

halibut_bank_winter <- summarise(group_by(halibut_sst_winter_grouped,
                                 winter_group),
                        year = max(year),
                        sst_winter = mean(sst)) %>%
  select(-c("winter_group"))

halibut_bank_winter
halibut_bank_winter %>% tail()

plot(halibut_bank_winter$year,
     halibut_bank_winter$sst_winter,
     type = "o",
     xlab = "Winter SST",
     ylab = "Year")
```


### Fraser River Discharge

- peak (max daily value) and average daily Fraser River Discharge from Apr-Jun
  (same as buoy temperatures). Had copied .rda files into data/ here, now just
  use pacea version.

So again, expect the lag to be two years.

Adapting temperature code from above, this is for mean discharge in spring:

```{r, fraser}
fraser_months <- 4:6

fraser_discharge_mean     # from pacea

fraser_mean_months <- dplyr::filter(fraser_discharge_mean,
                                    month %in% fraser_months)

# Check first year has all three months (i.e. starts in April);
fraser_mean_months    # Yes it does

fraser_mean_in_spring <- summarise(group_by(fraser_mean_months,
                                         year),
                        year = unique(year),
                        fraser_mean_spring = mean(value))

fraser_mean_in_spring
fraser_mean_in_spring %>% tail()

plot(fraser_mean_in_spring$year,
     fraser_mean_in_spring$fraser_mean_spring,
     type = "o",
     xlab = "Fraser River discharge in spring - mean",
     ylab = "Year")
```

And for peak discharge, which is the peak daily value over the spring months.

```{r, fraserpeak}
fraser_discharge_peak     # from pacea

fraser_peak_months <- dplyr::filter(fraser_discharge_peak,
                                    month %in% fraser_months)

# Check first year has all three months (i.e. starts in April);
fraser_peak_months    # Yes it does

fraser_peak_in_spring <- summarise(group_by(fraser_peak_months,
                                         year),
                        year = unique(year),
                        fraser_peak_spring = max(value))

fraser_peak_in_spring
fraser_peak_in_spring %>% tail()

plot(fraser_peak_in_spring$year,
     fraser_peak_in_spring$fraser_peak_spring,
     type = "o",
     xlab = "Fraser River discharge in spring - peak",
     ylab = "Year")
```

### Total Pink Salmon abundance in North Pacific

- total pink salmon abundance in N Pacific. Second or third years of marine
  life of the sockeye (as competition). Year after the temperature data (second
  year), so 2020 (for age4 returning in 2021). EDM should hopefully pick up on a
  lag of 1 year then (maybe 2 years also, for the second year of marine life of
  the sockeye).

  So the Pinks that are in the ocean in 2020.
  `np_salmon` has `return_year_of_np_salmon`. So we would want the
  `return_year_of_np_salmon` to be 2021, as those will be in the ocean in 2020.
  BUT, that implies `return_year = 2025` sockeye (what we want to predict), we will need the
  pinks at sea that are also returning in 2025, which we won't have.
  [Carrie contacting Hannah Hunter also].

The Pink Salmon only spend one year at sea. So can look at pinks returning in
2020 as interacting at sea in 2019 with the age4 Sockeye returning in 2021, when
the latter are age2. Ideally would want the pinks returning in 2021, but then to
predict Sockeye returning in 2025 we would need info on pinks returning in 2025,
which we do not have.

Now doing just the year here, meaning the `return_year_of_np_salmon`.

Have updated with Brendan's latest (10/6/26), 2024 is updated fromlast year's version, no 2025 yet
though; calling it `north_pacific_salmon`.

```{r npsalmon}
north_pacific_salmon

tail(north_pacific_salmon)

max(diff(north_pacific_salmon$return_year_of_np_salmon))   # No gaps
sum(is.na(north_pacific_salmon$total_ocean_salmon))                     # No missing values
```
Does start in 1925, which is nice. And fully continuous.

Adapting OISST code above, we just want to use the pinks and the total
salmon, which Brendan has already calculated, so do them together.

```{r, npsalmon2}
north_pacific_salmon_covariates <- rename(north_pacific_salmon,
                                          year = return_year_of_np_salmon) %>%
  select(-c("chum",
            "sockeye"))

north_pacific_salmon_covariates %>% tail()
```

### Total salmon abundance in N Pacific

- total salmon abundance in N Pacific. Same as pink. Andy: though won't the
  non-pinks be out there for longer so we should look at multiple years? Could
  do but will get complicated and have limited time, could always include lags
  in EDM analysis.

Done in previous section in the same way, for simplicity.

### Copepods the year of outmigration

Total biomass anomaly in SoG from pacea (same year as SST). So for 2021 returns we use 2019, so for
2025 we need 2023, which we do have. So again, hopefully EDM picks up a lag of 2
when we do everything by real year (not return year).

```{r zooplankton}
zooplankton_sog

zooplankton_sog %>% tail()

zoo <- select(zooplankton_sog,
              year,
              total_biomass) %>%
  rename(zoo_total_biomass = total_biomass)

zoo
tail(zoo)
```

### Bifurcation index the year of outmigration

Bifurcation index, year of outmigration, same year as SST (so same as zooplankton).
So, again, for 2021 returns we use 2019, so for 2025 we need 2023, which we do
have. Presumably for 2026 we need 2024, which we do have (2025 not been included yet).

Again, hopefully EDM picks up on the lag of 2.

```{r bi}
bi
bi %>% tail()

bi_input <- select(bi,
                   year,
                   anomaly) %>%
  rename(bi_anomaly = anomaly)

bi_input
tail(bi_input)
```

## Combine covariates into one tibble

```{r, combine}
fraser_covariates_2026 <- tibble(year = start_year:end_year) %>%
  left_join(pdo_winter) %>%
  left_join(halibut_bank_spring) %>%
  left_join(halibut_bank_winter) %>%  
  left_join(fraser_mean_in_spring) %>%
  left_join(fraser_peak_in_spring) %>%
  left_join(north_pacific_salmon_covariates) %>%
  left_join(zoo) %>%
  left_join(bi_input)

fraser_covariates_2026 %>% head() %>% a()
fraser_covariates_2026 %>% tail() %>% a()
```

User should then decide what years they require.

## Plots of covariates

Do a big crude plot of everything:
```{r plotcovariates, fig.height = 9}
to_plot <- tidyr::pivot_longer(fraser_covariates_2026,
                               cols = -year) %>%
  mutate(name = factor(name))

p <- ggplot(to_plot,
            aes(year, value)) +
  geom_path() +                        # leaves gaps for NA's, but none now
  # facet_wrap(~name, scales = "free_y")
  facet_grid(name ~ ., scales = "free_y")

p

# It reorders them alphabetically; need something like this:
#  mutate(name = factor(name, levels = paste0("X", 1:12))) %>%
```

That suggests, to get a long time series, let's first use the covariates that
span the full time series. [But now the x-axis is year, not return year.]

Full correlation plot (bit hard to see - see the later oen also)
```{r corrplot}
# pairs(fraser_covariates_2026,
#      lower.panel = panel.cor)
GGally::ggpairs(fraser_covariates_2026)
```

## Only covariates that span the full time series

Basically, which covariates have values in the first year. Then check they cover
all years. BUT have not determined what the first year is yet (since don't want
to pre-empt that for Kaitlyn, Steph etc. with their analyses).

So set the first year that we have age4's to be 1952 (as we did last year based
on the data). Note that for EDM analyses you will want to go back a few more
years to account for lags of the covariates impacting the age4's in 1952.

```{r fullspan}
age4_first_year <- 1952
age4_first_year_index <- which(fraser_covariates_2026$year == age4_first_year)
age4_first_year_index

cov_in_first_year <- which(!is.na(fraser_covariates_2026[age4_first_year_index, ]))

cov_in_first_year

names(fraser_covariates_2026)[cov_in_first_year]
```
So those are the time series that we have that start from `r age4_first_year`.

Pairs of those will definitely be correlated. But let's see:

```{r, fullspan2}
fraser_covariates_2026_span <- fraser_covariates_2026[ , cov_in_first_year]
summary(fraser_covariates_2026_span)    # Still some NA's.
fraser_covariates_2026_span %>% tail() %>% a()

GGally::ggpairs(fraser_covariates_2026_span)
```

A clearer plot (coefficients are calculated slightly differently):
```{r, fullspan3}
corrplot::corrplot(cor(fraser_covariates_2026_span,
                       use = "pairwise.complete.obs"),
  method = "number",
  type = "upper" # show only upper side
  )
```

In Haigh et al. (2019) we used a correlation coefficient between covaraties of
0.3 to exclude related covariates. Let's do that here. 

Just looking manually at the figure, we can see that should not have, in the
same analysis (because correlation has magnitude $>0.3$):

- `fraser_mean_spring` and `fraser_peak_spring` (as expected)

- `pink` and `total_ocean_salmon` (as expected)

- `pdo_winter_mean` and either `fraser_mean_spring` or `fraser_peak_spring`,
  though if we exclude `fraser_peak_spring` we should keep `pdo_winter_mean`;
  though fraser values are direct influences so should probably be preferred,
  but we get PDO values earlier I think.


## Save covariates

TODO In 2025 we removed the final row as I think it was messing up EDM because
it had some NA's. But do that in the analyses.

```{r, save}
usethis::use_data(fraser_covariates_2026,
                  overwrite = TRUE)   # Or maybe save your own copy as just an .rds. Change to
# TRUE if want to resave.
```

```{r, stop, cache = FALSE}
knitr::knit_exit()
```

Do not actually want the final year values, so remove those. Think it was
messing up the EDM. Was worth doing all years up to here, partly to see if we
can try and predict two years ahead (i.e. how fleshed out are recent covariates).


This is the final `input_age4`:
```{r, inputage4rev}
input_age4 <- input_age4_full[-nrow(input_age4_full), ]

input_age4 %>% head() %>% a()

input_age4 %>% tail() %>% a()

summary(input_age4)

T <- nrow(input_age4)   # T as in index T is final year of data, can include
                        # NA's at start. Not used in calculations, just plotting
                        # and printing results.
  #sum(!is.na(input_age4$age4)) # T in EDM speak, final index year of data, but
                               # now need to account for NA's at the start. So
                               # do it based on years of age4 values.
T    # do double check that manually, esp for age5 and other rivers

age4_first_year_index <- min(which(!is.na(input_age4$age4)))
age4_first_year_index

age4_first_year <- input_age4$year[age4_first_year_index]
age4_first_year
```

So only NA's in `sst_spring`, `zoo_total_biomass` and `bi_anomaly`, plus the
early ones for most variables. The ones in the tail look okay, given we know
what we expect the lags to be (don't allow MVE to use lags for which we don't
have the data for the 2025 prediction).


# Quick EDM, no covariates (text of later sections based on old inputs, needs updating)

Do a quick pbsEDM as proof of concept and to see what we get.

From `analyse_simple_time_series` vignette.

Try no covariates, iterate over $E$:

```{r, edm2}
res_univariate <- pbsEDM_Evec(input_age4$age4)
plot_pbsEDM_Evec(res_univariate,
                 portrait = FALSE)

plot_rho_Evec(res_univariate)
```
Does not have the three-armed structure like our simulated model did in our paper.

Those results suggest $E=5$ when using no covariates, and the prediction for 2025 would
be the last of these (I think):
```{r pred1}
res_univariate[[4]]$results

res_univariate[[4]]$N_forecast
res_univariate[[4]]$N_forecast[prediction_year_index]
```

Though the observed and forecast $N$ look quite different:
```{r obsN}
plot(res_univariate[[4]]$N_observed,
     res_univariate[[4]]$N_forecast)
```

and indeed, `N_rho$ is only `r round(res_univariate[[4]]$results$N_rho, 2)` as
seen in results above.

Checking the $X$:
```{r obsX}
plot(res_univariate[[4]]$X_observed,
     res_univariate[[4]]$X_forecast)
```

and indeed, `X_rho$ is only `r round(res_univariate[[4]]$results$X_rho, 2)` as
seen in the panel plot above.


\clearpage

# EDM with covariates

- Do a time series panel plot first of all time series.

- Then do the correlation plot first before truncating (expect it just ignores NA's),
and add in the correlation coefficient option.

- Now try some covariates, need to understand what is going on here. Lots of
options though. See multiview embedding also.

- But the method doesn't deal with NA's, and when keeping them in and doing the
full time series (starting from 1952) seems to just end up with a few years that
work; this was from doing chunk edm1 before removing NA's, which we're doing
here now.  Not sure about this now (in `analysis-2025-2.Rmd`).

```{r age4rmNA}
input_age4 %>% head() %>% a()
input_age4 %>% tail() %>% a()
summary(input_age4)
```
So we never want a lag of 0 for any covariates (only PDO would actually work for
making a prediction for 2025, though we've now taken out 2025 values as PDO was
the only one). Also, can't have lag of 1 for zooplankton, or lag of 2 for either spawners.


### Multivariate EDM for the full span values

Updating here, sporadically, based on top lag combination from mve.

(Earlier attempts at truncated time series are saved below, and used as template).

```{r edmspan1}
res_span1 <- pbsEDM(input_age4_span,
                    lags = list(age4 = c(0, 5),
                                # pdo_winter_mean = 0,
                                # fraser_mean_spring = 0,
                                pink = 1),
                    first_difference = TRUE,
                    centre_and_scale = TRUE)

res_span1$results
plot(res_span1, portrait = FALSE)  # technically not correct
```

This looked okay for multivariate. But multiview (below) it's better for EDM to
deal with the lags (and we have that automated already, and getting short of time).

$\rho$ here is very slightly different to version 2 (even though data haven't
changed - expect it's to do with the NA's, oh, and going back a bit further with
the pinks; though RMSE is the same). Don't worry too much.

Need to systematically go through the different options. Look at Luke's lag
function. Though we already have that working for multiview embedding, so let's
just use that. Can always try top lag choice back in multivariate.


### Multiview for simulated values

The following worked from `mve_understanding.Rmd` vignette in pbsEDM, to use as
a template:

It's all working okay. Try shebang one again with longer simulated data and more
lags. Though the actual top rho values shouldn't change from shebang; the values
are calculated in the function now.

```{r shebang2, eval = FALSE}
set.seed(42)
h_simulated <- 0.1095 + sample(1:180) * 0.001 # has mean of 0.2
simulated_4 <- EDMsimulate::salmon_sim(h = h_simulated)

lags_use_multi <- list(R_t = 0,
                       S_t = 0:3)

lags_lots <- list(R_t = 0:4,
                  S_t = 0:8)

mve_full_shebang_2 <- multiview_embedding(data = simulated_4,
                                          response = "R_t",
                                          lags = lags_lots)  # Do lags_use_multi
                                            # for shorter and quicker one (instantaneous);
                                            # lags_lots for longer, a few minutes
mve_full_shebang_2$rho_prediction_from_mve
mve_full_shebang_2$rho_each_top_subset
```

Wow, fascinating. The overall $\rho$ for mve, averaging over the top subsets, is
higher than every $\rho$ from every subset! So the averaging is indeed paying
off, and improving results. Other results are:

```{r shebang2res, eval = FALSE}
mve_full_shebang_2$lags_of_top_subsets  # These are in order of highest rho
mve_full_shebang_2$response_predicted_from_mve   # Overall mve prediction, including
                                            # for the forecast T+1.
```

Also, for calculating number of lag combinations to test:
Number of 5-dimensional combinations (?) is
```{r combinations}
choose(10, 1:10)
choose(10, 1:10) %>% sum()
choose(10, 1:10) %>% sum() %>% sqrt()
```
So that's manageable.

Should calculate $\rho$ only on what we care about, i.e. `R_t`.


### Multiview embedding for the span values

So, let's use multiview, and think about the lags. Can define the lags here. Since it
ignores other covariates, we don't need to create subsets of the data when
removing a correlated covariate.

Not sure if age4 lag of 0 should be in first line here - it
is for the example in pbsEDM, but then gets included in fitting. So not including it here,
then maybe trying. Okay, various different options:

For version 3 not changing these (even though we now have spawners also); still
don't see how we can use 0 lag of age4 though. Then adapting the bigger MVE analysis.

\clearpage


**1) No 0 lag of age4, 0 lag of PDO.**

Though not sure how it would use that in
prediction for 2025 (since haven't included that value now).

```{r, span1}
lags_span1 <- list(age4 = c(4, 5), # I guess 5 years back could relate to age5.
                   pdo_winter_mean = c(0, 1, 2), # expect 2
                   fraser_mean_spring = c(1, 2), # expect 2
                   pink = c(1, 2))

mve_span1 <- multiview_embedding(data = input_age4_span,
                                 response = "age4",
                                 lags = lags_span1)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

mve_span1$rho_prediction_from_mve
mve_span1$rho_each_top_subset

mve_span1$lags_of_top_subsets  # These are in order of highest rho
mve_span1$response_predicted_from_mve   # Overall mve prediction, including
                                            # for the forecast T+1.
```

```{r quickcompare}


```{r span1fig}
plot(input_age4_span$age4,
     mve_span1$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span1$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span1$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span1$response_predicted_from_mve[T+1], col = "red")
```

**2) No 0 lag of age4, no 0 lag of PDO.**

```{r, span2}
lags_span2 <- list(age4 = c(4, 5),     # I guess 5 years back could relate to
                                       # age5. Andy removed 0 lag here.
                   pdo_winter_mean = c(1, 2),   # expect 2
                   fraser_mean_spring = c(1, 2), # expect 2
                   pink = c(1, 2))               # expect 2, maybe 1 also.

mve_span2 <- multiview_embedding(data = input_age4_span,
                                 response = "age4",
                                 lags = lags_span2)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

mve_span2$rho_prediction_from_mve
mve_span2$rho_each_top_subset

mve_span2$lags_of_top_subsets  # These are in order of highest rho
mve_span2$response_predicted_from_mve   # Overall mve prediction, including
                                            # for the forecast T+1.
```

```{r span2fig}
plot(input_age4_span$age4,
     mve_span2$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span2$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span2$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span2$response_predicted_from_mve[T+1], col = "red")
```

<!-- Prediction for $T+1$ is `r f(mve_span2$response_predicted_from_mve[T+1])`. -->

**3) 0 lag of age4, 0 lag of PDO.**

Though not sure how it would use that in
prediction for 2025 (since haven't included that value now).

```{r, span3}
lags_span3 <- list(age4 = c(0, 4, 5), # I guess 5 years back could relate to age5.
                   pdo_winter_mean = c(0, 1, 2), # expect 2
                   fraser_mean_spring = c(1, 2), # expect 2
                   pink = c(1, 2))

mve_span3 <- multiview_embedding(data = input_age4_span,
                                 response = "age4",
                                 lags = lags_span3)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

mve_span3$rho_prediction_from_mve
mve_span3$rho_each_top_subset

mve_span3$lags_of_top_subsets  # These are in order of highest rho
mve_span3$response_predicted_from_mve   # Overall mve prediction, including
                                            # for the forecast T+1.
```

```{r span3fig}
plot(input_age4_span$age4,
     mve_span3$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span3$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span3$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span3$response_predicted_from_mve[T+1], col = "red")
```


$\rho$ is `r round(mve_span3$rho_prediction_from_mve, 2)` .

Prediction for $T+1$ is `r round(mve_span3$response_predicted_from_mve[T+1])`.

**4) 0 lag of age4, no 0 lag of PDO.**

Though not sure how it would use that in
prediction for 2025 (since haven't included that value now).

```{r, span4}
lags_span4 <- list(age4 = c(0, 4, 5), # I guess 5 years back could relate to age5.
                   pdo_winter_mean = c(1, 2), # expect 2
                   fraser_mean_spring = c(1, 2), # expect 2
                   pink = c(1, 2))

mve_span4 <- multiview_embedding(data = input_age4_span,
                                 response = "age4",
                                 lags = lags_span4)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

mve_span4$rho_prediction_from_mve
mve_span4$rho_each_top_subset

mve_span4$lags_of_top_subsets  # These are in order of highest rho
mve_span4$response_predicted_from_mve   # Overall mve prediction, including
                                            # for the forecast T+1.
```

```{r span4fig}
plot(input_age4_span$age4,
     mve_span4$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span4$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span4$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span4$response_predicted_from_mve[T+1], col = "red")
```



**5) Lots of lags of everything**

Give a lot of options (to run over lunch). Took a few hours but failed at the
.tex build, I think because it prints out lots and lots of lags. So not
rerunning but just showing the start of the results.

```{r, span5, eval = FALSE}
lags_span5 <- list(age4 = c(0, 1, 2, 3, 4, 5),
                   pdo_winter_mean = c(1, 2, 3, 4), # expect 2
                   fraser_mean_spring = c(1, 2, 3, 4), # expect 2
                   pink = c(1, 2, 3, 4))

mve_span5 <- multiview_embedding(data = input_age4_span,
                                 response = "age4",
                                 lags = lags_span5)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

mve_span5$rho_prediction_from_mve
mve_span5$rho_each_top_subset

mve_span5$lags_of_top_subsets  # These are in order of highest rho
mve_span5$response_predicted_from_mve   # Overall mve prediction, including for T+1.

saveRDS(mve_span5, file = "mve_span5.rds")
```

```{r, span5load, eval = FALSE}
mve_span5 <- readRDS("mve_span5.rds")

mve_span5$rho_prediction_from_mve
mve_span5$rho_each_top_subset

# So there's 512, so need to just show a few of these:
mve_span5$lags_of_top_subsets[1:20]

mve_span5$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

age4 with lag 0 is in the top 9 of those, and in 16 of the top 20. So should
retry without those.

```{r span5fig, eval = FALSE}
plot(input_age4_span$age4,
     mve_span5$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span5$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span5$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span5$response_predicted_from_mve[T+1], col = "red")
```

**6) Lots of lags of everything, except no 0 of age4 (or anything)**

Rerunning with NA's
removed when averaging over all SVE's (previous NA removal was the untransform
minimum replacement issue), so should have predictions for (probably)
all age4's (before we were excluding early years for which one of the SVEs did
not have all the data, as mean was not excluding NA's in `multiview_embedding()`.

```{r, span6}
lags_span6 <- list(age4 = c(1, 2, 3, 4, 5),
                   pdo_winter_mean = c(1, 2, 3, 4), # expect 2
                   fraser_mean_spring = c(1, 2, 3, 4), # expect 2
                   pink = c(1, 2, 3, 4))

mve_span6 <- multiview_embedding(data = input_age4_span,
                                 response = "age4",
                                 lags = lags_span6)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

mve_span6$rho_prediction_from_mve
mve_span6$rho_each_top_subset

mve_span6$lags_of_top_subsets  # These are in order of highest rho
mve_span6$response_predicted_from_mve   # Overall mve prediction, including for T+1.

saveRDS(mve_span6, file = "mve_span6.rds")
```

```{r, span6load}
mve_span6 <- readRDS("mve_span6.rds")

mve_span6$rho_prediction_from_mve
mve_span6$rho_each_top_subset

# So there's 512, so need to just show a few of these:
mve_span6$lags_of_top_subsets[1:20]

# Look at the top few predictions:

mve_span6[["response_each_subset"]][mve_span6$top_subsets_for_rho_index[1:5]]

mve_span6$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

Was true, but not updating span5 now since fixing NA issue in pbsEDM. Can actually see that top one here was the 10th one in span5:
```{r comparespan56, eval = FALSE}
mve_span6$rho_each_top_subset[1]
mve_span5$rho_each_top_subset[10]
```
which makes sense.

```{r span6fig}
plot(input_age4_span$age4,
     mve_span6$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span6$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span6$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span6$response_predicted_from_mve[T+1], col = "red")
```

**Just try span6 best model as multivariate EDM - not doing, keeping code**

Though need to include lag 0 also for the thing we're predicting.

```{r, span6multivariate, eval = FALSE, echo = FALSE}
mve_span6$lags_of_top_subsets[1]

multivariate_span6 <- pbsEDM(input_age4_span,
                      lags = list(age4 = c(0, 1, 3, 4),
                                  # pdo_winter_mean = 0,
                                  fraser_mean_spring = c(1, 4),
                                  pink = c(1, 3)),
                      first_difference = TRUE,
                      centre_and_scale = TRUE)

multivariate_span6$results
plot(multivariate_span6, portrait = FALSE)  # technically not correct
f(multivariate_span6$N_forecast[T+1])
```

$\rho$ based on $N$ came out lower than for span6, so stick with that. Worth
taking a look. Not sure why it's less than what is listed in span6 though. Ah,
it's only doing a few nearest neighbours, MVE does them all I think. Prediction
for $N$ is still low though, way under a million, similar to a couple of others.


**7) Lots of lags of everything, except no 0 of anything or lag 5 of age4 (since 1
and 5 came up in top five SVEs for span6 above**

Calcs took a while, so manually saving. (It failed when printing all the lags, which
I've fixed now).

```{r, span7}
if(FALSE){   # Want to redo chunk but not redo calc; tidy up when use as future
             # template for span8
  lags_span7 <- list(age4 = c(1, 2, 3, 4),
                     pdo_winter_mean = c(1, 2, 3, 4), # expect 2
                     fraser_mean_spring = c(1, 2, 3, 4), # expect 2
                     pink = c(1, 2, 3, 4))

  mve_span7 <- multiview_embedding(data = input_age4_span,
                                   response = "age4",
                                   lags = lags_span7)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()

  mve_span7$rho_prediction_from_mve
  mve_span7$rho_each_top_subset

  mve_span7$lags_of_top_subsets[1:20]  # These are in order of highest rho
  mve_span7$response_predicted_from_mve   # Overall mve prediction, including for T+1.

  saveRDS(mve_span7, file = "mve_span7.rds")
}

```

```{r, span7load}
mve_span7 <- readRDS("mve_span7.rds")

mve_span7$rho_prediction_from_mve
mve_span7$rho_each_top_subset

# So there's lots, so need to just show a few of these:
mve_span7$lags_of_top_subsets[1:20]

# Look at the top few predictions:

mve_span7[["response_each_subset"]][mve_span7$top_subsets_for_rho_index[1:5]]

mve_span7$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r span7fig}
plot(input_age4_span$age4,
     mve_span7$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_span7$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_span7$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_span7$response_predicted_from_mve[T+1], col = "red")
```



**8)Then try including other covariates that don't span full time series**

Want to include zooplankton. Just try it at first, and see if MVE just teases
anything out, ignoring years with no zooplankton data for the relevant SVEs
(giving NAs are now dealt with better in the code). Calling variables `full8`,
incrementing from `span7`. And add spawners, which I hadn't done yet.

Just trying a few to see if it works. Then copy back in similar lags to what was
used for span7.

Timings. Each extra lag should be $2^XX$ longer.

9 lags - 3.8 seconds

12 lags - 21 seconds (8 times longer, makes sense.

14 lags - 90 seconds (makes sense, four times longer, ish)

16 lags - should be about 320 seconds, 5 minutes.

17 lags - took 720 seconds, 12 minutes.

```{r, full8}
lags_full8 <- list(age4 = c(1, 2, 3, 4),
                   effective_spawners = c(4),
                   pdo_winter_mean = c(1, 2),
                   fraser_mean_spring = c(1, 2),
                   pink = c(1, 2),
                   sst_spring = c(1, 2),
                   zoo_total_biomass = c(1, 2),
                   bi_anomaly = c(1, 2))
tictoc::tic()
mve_full8 <- multiview_embedding(data = input_age4,   # not span
                                 response = "age4",
                                 lags = lags_full8)
# Centre and scaling and first differencing get done automatically, in
#  state_space_reconstruction_for_sve()
tictoc::toc()

saveRDS(mve_full8, file = "mve_full8.rds")
```

```{r, full8load}
mve_full8 <- readRDS("mve_full8.rds")

mve_full8$rho_prediction_from_mve
mve_full8$rho_each_top_subset

# So there's lots, so need to just show a few of these:
mve_full8$lags_of_top_subsets[1:20]

# Look at the top few predictions:

mve_full8[["response_each_subset"]][mve_full8$top_subsets_for_rho_index[1:5]]

mve_full8$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r full8fig}
plot(input_age4$age4,
     mve_full8$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_full8$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_full8$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_full8$response_predicted_from_mve[T+1], col = "red")
```

### Save covariates to use elsewhere

All this was just for Chilko. So save the non-Chilko-specific covariates to use
for other stocks (and age5 analysis).

```{r, saveinputage4}
fraser_covariates <- input_age4 %>%
  select(-c("age4",
            "effective_spawners",
            "total_spawners"))


```


```{r exit5, cache = FALSE}
knitr::knit_exit()
```


### Summary (needs updating)
UPDATE:

So far span6 MVE is best, has highest $\rho$ and does not have age4 (or anything)
with lag of 0, which we cannot use for prediction.

Interesting that all the predictions seem relatively low, though maybe they're extracting
the same main sets of lags? Bit hard to compare (a figure would be useful, or a
table).


### More notes

Can also see other methods in pbsEDM/R.

THere is more after this, just exiting to get running.

```{r exit2, cache = FALSE}
knitr::knit_exit()
```



### This was looking at the continuity of some of the time series, and doing a shorter timespan

So PDO is okay, need to look into others:
```{r age4rmNA2}
input_age4$sst_spring

input_age4$zoo_total_biomass

input_age4$bi_anomaly
```

So looks like it makes sense to start the analysis from the first year of
zooplankton data:
```{r age4rmNA3}
start_ind <- min(which(!is.na(input_age4$zoo_total_biomass)))
start_year <- input_age4$return_year[start_ind]
start_year

input_age4_truncated <- filter(input_age4,
                               return_year >= start_year)
input_age4_truncated
summary(input_age4_truncated)

plot(input_age4_truncated)
```

So now only have NA's in SST. Just ignore that data set for now and see what
multivariate EDM gives. So only 27 years of data, but 3 covariates, plus lags of
age4 (oh, but maybe exclude those, or include the lagged values we need somehow;
TODO maybe have to add a few years back for everything, and any NA's will get ignored).

The plot shows that `buoy_sst_otter` may indeed be having an influence (colder
is better). Gives motivation to back to the above, and quickly go through SST
for the third buoy, East Dellwood Knolls, in case that looks better and has less
NA's. No it doesn't, so now going back up to Otter buoy and just replacing the
few NA's we need with means of adjacent years; not ideal, but a simple fix.


This analysis does not include the SST (since have to deal with the NA's, and
first going back up to look at the East Dellwood Knolls buoy).
```{r edm1}
res <- pbsEDM(input_age4_truncated,
              lags = list(age4 = c(0, 4, 5),
                          pdo_winter_mean = 0,
                          # buoy_sst_otter = 0,
                          zoo_total_biomass = 0,
                          bi_anomaly = 0),
              first_difference = TRUE,
              centre_and_scale = TRUE)

res$results
plot(res, portrait = FALSE)  # technically not correct
```

So $\rho$ is actually negative :o

Try no lags of age4:
```{r edm2a}
res2 <- pbsEDM(input_age4_truncated,
               lags = list(age4 = 0,
                          pdo_winter_mean = 0,
                          # buoy_sst_otter; = 0,
                          zoo_total_biomass = 0,
                          bi_anomaly = 0),
              first_difference = TRUE,
              centre_and_scale = TRUE)

res2$results
# plot(res2, portrait = FALSE)  # technically not correct, though only the time
                              # series I think
plot(res2$N_observed, res2$N_forecast)
```

So $\rho$ is even worse.


```{r exit, cache = FALSE}
knitr::knit_exit()
```

## Multiview embedding

From `mve_understanding` sort-of-vignette:
```{r mve}
lags_for_mve <- list(age4 = 1:5,   # 0 gives an error, whih makes sense as the response
                     at_sea_abundance = 0,
                     pdo_winter_mean = 0)
mve_full <- multiview_embedding(data = input_age4,
                                response = "age4",
                                lags = lags_for_mve)

# These look to be useful:

mve_full$top_subsets_for_rho_index
mve_full$rho_each_top_subset
mve_full$lags_of_top_subsets

mve_full$response_predicted_from_each_top_subset %>% tail()  # Note that row 74,
                                        # prediction_year_index, has two the
                                        # same (for age4=1:2 in lags).

mve_full$rho_prediction_from_mve

mve_full$response_predicted_from_mve

mve_full$response_predicted_from_mve[prediction_year_index]  # The year of interest

```

Last one has NA from just one prediction. It's this:
`
[[9]]$at_sea_abundance
[1] 0

[[9]]$pdo_winter_mean
[1] 0
`
so excluding any age 4. Should still be able to give a prediction though.

Could do this, but think it's meant to be weighted. Have to dig into the code:
```{r mean}
mean(mve_full$response_predicted_from_each_top_subset[prediction_year_index,],
     na.rm = TRUE)
```
Plus there are lots of NA's in earlier predictions also, so need to figure that out.

Look into how it picks the top embeddings.
