---
title: "Fraser River 2026 - single stock analysis"
author: "Andrew Edwards and Carrie Holt"
output: pdf_document
fontsize: 12pt
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  cache = TRUE,
  cache_path = "fraser-analysis-cache/",
  fig.path = "fraser-analysis-figs-cache/",
  fig.width = 7,
  fig.height = 6
)

knitr::dep_prev()
```

```{r, build, echo = FALSE, eval = FALSE}
# To build either run this line or click knit button in RStudio:
rmarkdown::render("fraser-analysis.Rmd")
```

```{r, packages, echo = FALSE, cache = FALSE}
load_all()
# Or:
# remotes::install_github("andrew-edwards/sockeyePrize")
# library(sockeyePrize)
library(dplyr)
library(pbsEDM)
library(pacea)    # make sure to update
library(gridExtra)
library(ggplot2)
```

```{r, river}
river <- "Chilko"
```

Adapting from 2025 code, having saved `fraser_covariates_2026` in this
package. Using Columbia analysis that I did first as it was the first one we got
data for for the competition.

Getting this running, then TODO checking the covariates we used last year for
Fraser and switch to using those.

TODO also check the ages we did last year for Fraser. See
`report/fraser/late_stuart_simplified/analysis-2025-4-simplified.Rmd`.


TODO From chat with Kaitlyn and Jessie, SST over winter is indicative of food
available, and I think has been more hightly ranked in some of their
analyses. Better to use Dec-Feb or rather Nov-Mar, so as a proxy for other parts
of the ecosystem. So check what I did, and then update as necessary. Actually we
did Apr-Jul so quite different.

This is for age 3s, 4s, and 5s for `r river`, though have to do four calculations.

If you changed the lags to rerun the multiview embedding, you need to delete the
saved .rds file.

# Fraser River

What we would put into pbsEDM.

## River-specific salmon data

Try to predict the following (data are saved in `returns_fraser_2026`):

- `age11` (my notation): AgeClass_1.1 returns each year, **say 2021 as an example**. 
  
- `age12`: AgeClass_1.2 returns each year, (2021), an additional covariate would be AgeClass_1.1 returns.

- `age13`: AgeClass_1.3 returns each year, (2021), an additional covariate would be AgeClass_1.2 returns.

- `age22`: AgeClass_2.2 returns each year, (2021)

Do each separately and then give the sum as the total prediction.

The full data are:

```{r, returns}
returns_fraser_2026
summary(returns_fraser_2026)
```

Later we will extract values for just the one river: `r river`


## Environmental covariates

Covariates would be (with their names in `fraser_covariates_2026` data tibble
constructed in `fraser-covariates-2026.Rmd`):

- PDO in the year prior to outmigration (Nov 2019-March 2020) for returning
  AgeClass_1.1. (`pdo_winter_mean`)

- PDO in the year prior to outmigration (Nov 2018-March 2019) for returning AgeClass_1.2 and AgeClass_2.2.

- PDO in the year prior to outmigration (Nov 2017-March 2018) for returning AgeClass_1.3.

- As for Fraser River, was going to get temperature from buoys: "Halibut Bank" (Apr-Jun average),
  "West Sea Otter" (Apr-Jul average), "East Dellwood Knolls" (Apr-Jul average),
  Year of outmigration, after the PDO Mar calculation. SOme missing data for
  those I think, which is why went for OISST for the pixel closest to Halibut
  Bank buoy, then averaged each year over April to July. Saved as column `sst_spring`.

- total pink salmon abundance in N Pacific returning the same return year as
  being predicted (2021), to predict AgeClass 1.1, AgeClass_1.2 and AgeClass_2.2
  returns (noting pink salmon might not compete with AgeClass_1.1). Column is `pink`.

- total pink salmon abundance in N Pacific returning the year prior to the year
  being predicted (2020), to predict AgeClass_1.3 returns (an alternative
  hypothesis is that pink salmon in the same return year as being predicted
  could be used for this age class as well). Column is `total_ocean_salmon`.

- copepods the year of outmigration (same as SST). Total biomass anomaly. Might
  not have such a long time series; `zoo_total_biomass`.

- bifurcation index, year of outmigration. `bi_anomaly`.

# Returns

See TODOs above.

Not doing combined age4 then age5 like originally did for Fraser River, rather doing each of
the four listed above. Think can just make one big tibble with everything, and
then select the response variable appropriately.

```{r, stop, cache = FALSE}
knitr::knit_exit()
```


Just want the four age classes listed above (we did include more for Fraser River), and rename them:

```{r classes}
classes <- c("AgeClass_1.1",
             "AgeClass_1.2",
             "AgeClass_1.3",
             "AgeClass_2.2")

river_returns <-
  dplyr::filter(returns_fraser_2026,
                River == river) %>%
  dplyr::select(ReturnYear,
                Total_Returns,
                unlist(classes)) %>%
  rowwise() %>%
  mutate(total_of_included = sum(c_across(any_of(classes)),
                                 na.rm = TRUE)) %>%
  mutate(total_difference = Total_Returns - total_of_included) %>% ungroup()
# to check what we're missing by excluding some classes

river_returns
summary(river_returns)
summary(river_returns$total_of_included)
summary(river_returns$total_difference)
select(river_returns, ReturnYear, total_difference)
max(river_returns$total_of_included)
max(river_returns$total_difference)
```
So max excluded is about `r
f(max(river_returns$total_difference)/max(river_returns$total_of_included)* 100)`% of the maximum total (might not be the same year),
which is fine.

```{r rename}
input_returns <- select(river_returns,
                        year = ReturnYear,
                        age11 = AgeClass_1.1,
                        age12 = AgeClass_1.2,
                        age13 = AgeClass_1.3,
                        age22 = AgeClass_2.2)
summary(input_returns)

input_returns %>% a()
```

Just one actual 0. No need to replace any leading zeros
like for Fraser River.

```{r, back}
# Now add five earlier years with NA's, since things like PDO and spawners might
#  go back 4-5 years.
yr <- min(input_returns$year)
yr
input_returns <- rbind(c(yr - 5, rep(NA, 4)),
                     c(yr - 4, rep(NA, 4)),
                     c(yr - 3, rep(NA, 4)),
                     c(yr - 2, rep(NA, 4)),
                     c(yr - 1, rep(NA, 4)),
                     input_returns)
input_returns
input_returns %>% tail()

# For Columbia, not sure if needed:
if(min(input_returns$year != 1980)){
  stop("check covariates and years")
}

```

## Columbia covariates

Now join with the covariates, but remove Fraser River spring flows:

```{r, returnscov}
input_returns_covariates <- left_join(input_returns,
                                      fraser_covariates_2026) %>%
  select(-c("fraser_mean_spring",
            "fraser_peak_spring"))
input_returns_covariates
head(input_returns_covariates) %>% a()

tail(input_returns_covariates) %>% a()
```

Prob better to automate from the results instead of this TODO:
```{r, Tcalc}
T <- nrow(input_returns_covariates)   # T as in index T is final year of data, can include
                        # NA's at start. Not used in calculations, just plotting
                        # and printing results.

T    # do double check that manually, esp for age5 and other rivers
```

# EDM with covariates

Do a big plot of everything:
```{r plotcovariates, fig.height = 9}
to_plot <- tidyr::pivot_longer(input_returns_covariates,
                               cols = -year) %>%
  mutate(name = factor(name))

p <- ggplot(to_plot,
            aes(year, value)) +
  geom_path() +                        # leaves gaps for NA's, but none now
  # facet_wrap(~name, scales = "free_y")
  facet_grid(name ~ ., scales = "free_y")

p

# It reorders them alphabetically; need something like this:
#  mutate(name = factor(name, levels = paste0("X", 1:12))) %>%
```

```{r corrplot}
# pairs(input_returns_covariates,
#      lower.panel = panel.cor)
GGally::ggpairs(input_returns_covariates)
```

Useful, but of course doesn't highlight correlation of lags.
```{r corrplot22, fig.width = 9}
corrplot::corrplot(cor(input_returns_covariates,
                       use = "pairwise.complete.obs"),
  method = "number",
  type = "upper" # show only upper side
  )
```

Go with removing environmental variables that have correlations $>0.3$ (based on
what we did in Haigh et al.), just do manually based on this plot:
```{r corrplot23, fig.width = 9}
corrplot::corrplot(cor(select(input_returns_covariates,
                              -(year:age22)),
                       use = "pairwise.complete.obs"),
  method = "number",
  type = "upper" # show only upper side
  )
```

Coefficients $>0.3$ and decision (just keep `pink` like last year, as `pink`
dominate the `total_ocean_salmon`; though maybe more logical to keep total, so
ignore `total_ocean_salmon`?)

- `pdo_winter_mean` and `pink` - quite distinct variables, so keep both.

- `pdo_winter_mean` and `bi_anomaly` - `bi` is shorter so remove, ignore in
  remainder of this list

- `sst_spring` and `pink` - again, distinct, so keep both

Using this in calls below to `multiview_embedding()`.

# MVE analysis

Age12 are aged 4!!! etc.

For Fraser, for age4 we did age4 lags of `c(1, 2, 3, 4)`, and for age5 added lag
of 5. So take same approach here, similarly for age12 since age 3. Plus two
years of lags for age4 environmental covariates, and the previous returns as
outlined above.

## Age11

TODO may need to remove final row as has NA's and think that caused issues
before.

These are age3 when returning. Spend a whole year in freshwater first.

```{r, age11calc}
age11_lags <- list(age11 = c(1, 2, 3),
                   pdo_winter_mean = c(1),
                   pink = c(1, 2),
                   sst_spring = c(1))
if(!file.exists("age11_res.rds")){
  tictoc::tic()
  age11_res <- multiview_embedding(data = input_returns_covariates,
                                   response = "age11",
                                   lags = age11_lags)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()
  tictoc::toc()

  saveRDS(age11_res, file = "age11_res.rds")
} else {
  age11_res <- readRDS("age11_res.rds")
}

```

Results:

```{r, age11load}
age11_res$rho_prediction_from_mve
age11_res$rho_each_top_subset

# So there's lots, so need to just show a few of these:
age11_res$lags_of_top_subsets

# Look at the top few predictions:

age11_res[["response_each_subset"]][age11_res$top_subsets_for_rho_index[1:5]]

age11_res$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r age11fig}
plot(input_returns_covariates$age11,
     age11_res$response_predicted_from_mve[-(T+1)],
     xlab = "Age11 returns - observed",
     ylab = "Age11 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(age11_res$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(age11_res$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = age11_res$response_predicted_from_mve[T+1], col = "red")
```

### Age12

Four year olds when return.
```{r, age12calc}
age12_lags <- list(age12 = c(1, 2, 3, 4),
                   age11 = c(1),
                   pdo_winter_mean = c(1, 2),
                   pink = c(1, 2),
                   sst_spring = c(1, 2))

if(!file.exists("age12_res.rds")){
  tictoc::tic()
  age12_res <- multiview_embedding(data = input_returns_covariates,
                                   response = "age12",
                                   lags = age12_lags)
  # Centre and scaling and first differencing get done automatically, in
#  state_space_reconstruction_for_sve()
  tictoc::toc()

  saveRDS(age12_res, file = "age12_res.rds")
} else {
  age12_res <- readRDS("age12_res.rds")
}
```

Results:

```{r, resultsage12}
age12_res$rho_prediction_from_mve
age12_res$rho_each_top_subset

# So there's lots, so need to just show a few of these:
age12_res$lags_of_top_subsets[1:10]

# Look at the top few predictions:

age12_res[["response_each_subset"]][age12_res$top_subsets_for_rho_index[1:5]]

age12_res$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r age12fig}
plot(input_returns_covariates$age12,
     age12_res$response_predicted_from_mve[-(T+1)],
     xlab = "Age12 returns - observed",
     ylab = "Age12 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(age12_res$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(age12_res$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = age12_res$response_predicted_from_mve[T+1], col = "red")
```

# Age13

Five year olds, could add age11 but there are not many of them.

```{r, age13calc}
age13_lags <- list(age13 = c(1, 2, 3, 4, 5),
                   age12 = c(1),
                   pdo_winter_mean = c(1, 2, 3),
                   pink = c(1, 2, 3),
                   sst_spring = c(1, 2, 3))

if(!file.exists("age13_res.rds")){
  tictoc::tic()
  age13_res <- multiview_embedding(data = input_returns_covariates,
                                   response = "age13",
                                   lags = age13_lags)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()
  tictoc::toc()
  
  saveRDS(age13_res, file = "age13_res.rds")
} else {
  age13_res <- readRDS("age13_res.rds")
}
```

Results
```{r, age13res}
age13_res$rho_prediction_from_mve
age13_res$rho_each_top_subset

# So there's lots, so need to just show a few of these:
age13_res$lags_of_top_subsets[1:10]   

# Look at the top few predictions:

age13_res[["response_each_subset"]][age13_res$top_subsets_for_rho_index[1:5]]

age13_res$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r age13fig}
plot(input_returns_covariates$age13,
     age13_res$response_predicted_from_mve[-(T+1)],
     xlab = "Age13 returns - observed",
     ylab = "Age13 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(age13_res$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(age13_res$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = age13_res$response_predicted_from_mve[T+1], col = "red")
```

# Age22

Five year olds, could add age11 but there are not many of them.

```{r, age22calc}
age22_lags <- list(age22 = c(1, 2, 3, 4, 5),
                   pdo_winter_mean = c(1, 2, 3),
                   pink = c(1, 2, 3),
                   sst_spring = c(1, 2, 3))

if(!file.exists("age22_res.rds")){
  tictoc::tic()
  age22_res <- multiview_embedding(data = input_returns_covariates,
                                   response = "age22",
                                 lags = age22_lags)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()
  tictoc::toc()

  saveRDS(age22_res, file = "age22_res.rds")
} else {
  age22_res <- readRDS("age22_res.rds")
}
```
Results

```{r, age22res}
age22_res$rho_prediction_from_mve
age22_res$rho_each_top_subset

# So there's lots, so need to just show a few of these:
age22_res$lags_of_top_subsets[1:10]

# Look at the top few predictions:

age22_res[["response_each_subset"]][age22_res$top_subsets_for_rho_index[1:5]]

age22_res$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r age22fig}
plot(input_returns_covariates$age22,
     age22_res$response_predicted_from_mve[-(T+1)],
     xlab = "Age22 returns - observed",
     ylab = "Age22 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(age22_res$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(age22_res$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = age22_res$response_predicted_from_mve[T+1], col = "red")
```

## Summarise results

Summarise results of the four different age classes, and add together for final prediction.

```{r, summarylim}
lim_common_max <- max(c(input_returns_covariates$age11,
                        input_returns_covariates$age12,
                        input_returns_covariates$age13,
                        input_returns_covariates$age22,
                        age11_res$response_predicted_from_mve,
                        age12_res$response_predicted_from_mve,
                        age13_res$response_predicted_from_mve,
                        age22_res$response_predicted_from_mve),
                      na.rm = TRUE)
lim_common_max
lim_common = c(0, lim_common_max)
```

Do panel plot of time series, not repeating the earlier ones.
```{r, summaryplot, fig.height = 9}
par(mfrow = c(4,1))

# Age11 time series
plot(input_returns_covariates$year,
     input_returns_covariates$age11,
     type = "o",
     main = "Age11",
     xlab = "Year",
     ylab = "Age11 returns",
     ylim = lim_common)

points(c(input_returns_covariates$year, max(input_returns_covariates$year) + 1),
       age11_res$response_predicted_from_mve,
       pch = 16,
       col = "red",
       cex = 0.8)

legend("topleft",
       legend = c("observed", "predicted"),
       pch = c(1, 16),
       col = c("black", "red"))

# Age12 time series
plot(input_returns_covariates$year,
     input_returns_covariates$age12,
     type = "o",
     main = "Age12",
     xlab = "Year",
     ylab = "Age12 returns",
     ylim = lim_common)

points(c(input_returns_covariates$year, max(input_returns_covariates$year) + 1),
       age12_res$response_predicted_from_mve,
       pch = 16,
       col = "red",
       cex = 0.8)

legend("topleft",
       legend = c("observed", "predicted"),
       pch = c(1, 16),
       col = c("black", "red"))

# Age13 time series
plot(input_returns_covariates$year,
     input_returns_covariates$age13,
     type = "o",
     main = "Age13",
     xlab = "Year",
     ylab = "Age13 returns",
     ylim = lim_common)

points(c(input_returns_covariates$year, max(input_returns_covariates$year) + 1),
       age13_res$response_predicted_from_mve,
       pch = 16,
       col = "red",
       cex = 0.8)

legend("topleft",
       legend = c("observed", "predicted"),
       pch = c(1, 16),
       col = c("black", "red"))


# Age22 time series
plot(input_returns_covariates$year,
     input_returns_covariates$age22,
     type = "o",
     main = "Age22",
     xlab = "Year",
     ylab = "Age22 returns",
     ylim = lim_common)

points(c(input_returns_covariates$year, max(input_returns_covariates$year) + 1),
       age22_res$response_predicted_from_mve,
       pch = 16,
       col = "red",
       cex = 0.8)

legend("topleft",
       legend = c("observed", "predicted"),
       pch = c(1, 16),
       col = c("black", "red"))
```

Summarise, and then sum together age4 and age5 to get our prediction.

```{r, summarise}
summary <- tibble(class = "age11",
                  rho = age11_res$rho_prediction_from_mve ,
                  prediction = age11_res$response_predicted_from_mve[T+1])
summary <- rbind(summary,
                 c("age12",
                   age12_res$rho_prediction_from_mve,
                   age12_res$response_predicted_from_mve[T+1]),
                 c("age13",
                   age13_res$rho_prediction_from_mve,
                   age13_res$response_predicted_from_mve[T+1]),
                 c("age22",
                   age22_res$rho_prediction_from_mve,
                   age22_res$response_predicted_from_mve[T+1])) %>%
  dplyr::mutate_at(vars(class),
                   factor) %>%
  dplyr::mutate_at(vars(rho,
                        prediction),
                   as.numeric)

summary

knitr::kable(summary,
             digits = c(NA, 2, 6))
stopifnot(nrow(summary) == 4)

total_returns_prediction <- sum(summary$prediction)
```

The prediction for combined age4 and age5 returns for `r river` is ... `r f(total_returns_prediction)`.
