---
title: "Fraser River 2026 - single stock analysis"
author: "Andrew Edwards and Carrie Holt"
output: pdf_document
fontsize: 12pt
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  cache = TRUE,
  cache_path = "fraser-analysis-cache/",
  fig.path = "fraser-analysis-figs-cache/",
  fig.width = 7,
  fig.height = 6
)

knitr::dep_prev()
```

```{r, build, echo = FALSE, eval = FALSE}
# To build either run this line or click knit button in RStudio:
rmarkdown::render("fraser-analysis.Rmd")
```

```{r, packages, echo = FALSE, cache = FALSE}
load_all()
# Or:
# remotes::install_github("andrew-edwards/sockeyePrize")
# library(sockeyePrize)
library(dplyr)
library(pbsEDM)
library(pacea)    # make sure to update
library(gridExtra)
library(ggplot2)
```

```{r, river}
river <- "Chilko"
```

Getting this running, then doing the covariates we used last year for
Fraser and switch to using those.

Using the ages we did last year for Fraser. See
`report/fraser/late_stuart_simplified/analysis-2025-4-simplified.Rmd`.

We did all age4 and then all age5 and added, whereas for Columbia we did age11,
age12, age13, age22. I think the difference is because there would be too many
for FR, or the smaller age classes are less important, but surely the lags are
out of sync with the actual influences. TODO Chat further with Carrie and others
one day.

Done. From chat with Kaitlyn and Jessie, SST over winter is indicative of food
available, and I think has been more hightly ranked in some of their
analyses. Better to use Dec-Feb or rather Nov-Mar, so as a proxy for other parts
of the ecosystem. So check what I did, and then update as necessary. Actually we
did Apr-Jul so quite different.

For Columbia did:  age 3s, 4s, and 5s for `r river`, though have to do four
calculations. FOr FR 2025 looks like we combined data and did age4 and age5,
hence doing that here.

If you changed the lags to rerun the multiview embedding, you need to delete the
saved .rds file.

## Environmental covariates

Example years are for **returns in 2021 as an example**.

Covariates would be (with their names in `fraser_covariates_2026` data tibble
constructed in `fraser-covariates-2026.Rmd`):

- PDO in the year prior to outmigration (Nov 2019-March 2020) for returning
  AgeClass_1.1. (`pdo_winter_mean`)

- PDO in the year prior to outmigration (Nov 2018-March 2019) for returning AgeClass_1.2 and AgeClass_2.2.

- PDO in the year prior to outmigration (Nov 2017-March 2018) for returning AgeClass_1.3.

- Was going to get temperature from buoys: "Halibut Bank" (Apr-Jun average),
  "West Sea Otter" (Apr-Jul average), "East Dellwood Knolls" (Apr-Jul average),
  Year of outmigration, after the PDO Mar calculation. SOme missing data for
  those I think, which is why went for OISST for the pixel closest to Halibut
  Bank buoy, then averaged each year over April to July. Saved as column `sst_spring`.

- Kaitlyn and Jessie suggested winter temp from Nov-Mar, so doing that as
  `sst_winter`, the year of outmigration (so same as `sst_spring`).

- total pink salmon abundance in N Pacific returning the same return year as
  being predicted (2021), to predict AgeClass 1.1, AgeClass_1.2 and AgeClass_2.2
  returns (noting pink salmon might not compete with AgeClass_1.1). Column is `pink`.

- total pink salmon abundance in N Pacific returning the year prior to the year
  being predicted (2020), to predict AgeClass_1.3 returns (an alternative
  hypothesis is that pink salmon in the same return year as being predicted
  could be used for this age class as well). Column is `total_ocean_salmon`.

- copepods the year of outmigration (same as SST). Total biomass anomaly. Might
  not have such a long time series; `zoo_total_biomass`.

- bifurcation index, year of outmigration. `bi_anomaly`.

# Fraser River

What we would put into pbsEDM.

This is from 2025
`fraser/late_stuart_simplified/analysis-2025-4-simplified.Rmd`, then combine
some of what I have after exit chunk from Columbia, maybe.


This is for age4 and age5 for `r river`. The main multiview embedding can take a
while to run, see the variables `rerun_age4` and `rerun_age5` later and set them
to `FALSE` after you have run the code once.

# Returns

```{r classes, eval = FALSE}
# 2025 code, not evaluating:
classes_age4 <- c("AgeClass_0.3",
                  "AgeClass_1.2",
                  "AgeClass_2.1")

classes_age5 <- c("AgeClass_0.4",
                  "AgeClass_1.3",
                  "AgeClass_2.2",
                  "AgeClass_3.1")
returns_fraser_2026
summary(returns_fraser_2026)
```

I got confused in 2026 about the above, since we are looking at things like PDO
in the year prior to outmigration, which will not be the same for all the
age4's. Should really do each age class separately, but no time for that. Carrie
said: I think for 2025, we assumed almost all age fours were 1.2's and all age fives were 1.3's (which is pretty close to correct), and made the alignments accordingly. 

So seems more logical to just do the 1.2's for age4 and 1.3's for age5, so
thought about doing
that for 2026, except 2.2 has lots of 0's but also some big values, and 1.3 has
some NA's and some 0's. So let's use both for age5:

```{r classes2026}
# 2026 code:
returns_fraser_2026
summary(returns_fraser_2026)

classes_age4 <- c(# "AgeClass_0.3",
                  "AgeClass_1.2")
                  #"AgeClass_2.1")

classes_age5 <- c(# "AgeClass_0.4",
                  "AgeClass_1.3",
                  "AgeClass_2.2")
                  # "AgeClass_3.1")   # Seem to be none anyway
```

We want to predict `returns_fraser_2026$Total_Returns`, by doing age4 and age5
separately and then sum.

```{r sumreturns}
one_returns <- dplyr::filter(returns_fraser_2026,
                             River == river)

one_returns_age45 <- rowwise(one_returns) %>%
  mutate(age4 = sum(c_across(any_of(classes_age4)),
                    na.rm = TRUE)) %>%
  mutate(age5 = sum(c_across(any_of(classes_age5)),
                    na.rm = TRUE)) %>%
  ungroup() %>%
  select("System",
         "River",
         "ReturnYear",
         "age4",
         "age5")

one_returns_age45 %>% a()

input_returns <- select(one_returns_age45,      # only for Chilko
                        year = ReturnYear,
                        age4,
                        age5) %>%
  filter(year > 1951)       # since age4 and age5 are 0 for 1950 (newly added for 2026!) and 1951.

# Carrie said to keep the zeros in for Late Stuart 1952 and 1953 (there are some
# later as well, which the following code would have removed); these probably won't be predicted by MVE since no lags, but could get
# used as nearest neighbours.
# which(input_returns == 0)    # 1952 age5 for Chilko, check others manually here
# input_returns[input_returns == 0] = NA

input_returns

# in analysis-2025-3.Rmd had added NA for 2025, but then take out later, so just
# not bothering here.

# Now add five earlier years with NA's, since things like PDO and spawners might
#  go back 4-5 years.
yr <- min(input_returns$year)
yr
input_returns <- rbind(c(yr - 5, NA, NA),
                     c(yr - 4, NA, NA),
                     c(yr - 3, NA, NA),
                     c(yr - 2, NA, NA),
                     c(yr - 1, NA, NA),
                     input_returns)
input_returns
input_returns %>% tail()

if(min(input_returns$year != 1947)){
  stop("check covariates and years")
}

```

## Spawners

For Chilko and other FR, we want to use effective spawners (which we don't have
for other systems). Brood year is in there, which is what we want to use
and let EDM do the lagging.

They changed effective spawners set up. I just saved in one data object, think
can just join with the correct river and then use the correct lag for age4 or age4.

```{r, eff1}
effective_spawners_2026
# This does not work this year, since column names have changed
#eff <- select(effective_spawners_2026,
#              River,
#              Euro,
#              BroodYear,
#              Total_Spawners_BroodYear,
#              Total_EFS_BroodYear)
#eff
#summary(eff)
```

think can ignore this chunk in 2026:


```{r, onespanw, eval = FALSE}
one_spawners <- filter(eff_summ,
                       River == river)

one_spawners  # Great, it starts four years before first returns, so might mess
              # up for EDM for age5, but EDM will take care of it?

one_spawners_use <- select(one_spawners,
                           year = BroodYear,
                           total_spawners = total_spawners_brood_year,
                           effective_spawners = efs_brood_year)
one_spawners_use %>% a()
```

Now join returns, spawners, and System-wide covariates into one tibble.

```{r input}
effective_spawners_river <- select(effective_spawners_2026,
                                   year = BroodYear,   # year is return year in
                                   # analysis, so need to specify the correct
                                   # lag for EDM
                                   effective_spawners = all_of(river))

effective_spawners_river

input_river <- left_join(input_returns,
                         effective_spawners_river) %>%
  left_join(fraser_covariates_2026)
input_river %>% head(10) %>% a()
input_river %>% tail() %>% a()
```

```{r, indices}
summary(input_river)

T <- nrow(input_river)   # T as in index T is final year of data, can include
                        # NA's at start. Not used in calculations, just plotting
                        # and printing results.

T    # do double check that manually, esp for age5 and other rivers

age4_first_year_index <- min(which(!is.na(input_river$age4))) # generalise for age4
#and age5 if needed
age4_first_year_index

age4_first_year <- input_river$year[age4_first_year_index]
age4_first_year
```

# EDM with covariates

Do a big plot of everything:
```{r plotcovariates, fig.height = 9}
to_plot <- tidyr::pivot_longer(input_river,
                               cols = -year) %>%
  mutate(name = factor(name))

p <- ggplot(to_plot,
            aes(year, value)) +
  geom_path() +                        # leaves gaps for NA's, but none now
  # facet_wrap(~name, scales = "free_y")
  facet_grid(name ~ ., scales = "free_y")

p

# It reorders them alphabetically; need something like this:
#  mutate(name = factor(name, levels = paste0("X", 1:12))) %>%
```

Not worrying about span idea for now, can select the relevant covariates as lags
in EDM (i.e. just exclude ones that don't have full span).

```{r corrplot}
# pairs(input_river,
#      lower.panel = panel.cor)
GGally::ggpairs(input_river)
```

Useful, but of course doesn't highlight correlation of lags.
```{r corrplot22, fig.width = 9}
corrplot::corrplot(cor(input_river,
                       use = "pairwise.complete.obs"),
  method = "number",
  type = "upper" # show only upper side
  )
```
 


Using 0.3 to exclude related covariates. Updating these notes for stock-specific
covariates:

- `total_spawners` (don't have wrangled in 2026) and `effective_spawners` (as expected), go with effective (Catarina).

- Chilko: `total_spawners` or `effective_spawners` with either `pink` or `total_ocean_salmon`; though
  could always try (0.3 cut off is a bit arbitrary). I think we should stick
  with `effective_spawners` (if available) and `pink` though (even though 0.42 correlation). Late Stuart: TODO 2025: not correlated now
  (so bit more justified in sticking with them). Very cyclic compared to Chilko.

So, as for Chilko 2025, remove:

- `fraser_peak_spring`

- `total_ocean_salmon`

- `total_spawners`

- `bi_anomaly`

Let's try keeping `sst_spring` and `sst_winter` to see what comes out. 

2025: Given BI was not in the best runs for either age4 or age5 for Chilko, not going
to bother with it going forward (and correlated with PDO anyway).


# MVE analysis

Not doing an 'everything as baseline' run either, since takes a while. And just
including PDO (pdo might still appear in some variable names, as for Chilko we
tried PDO and BI separately).

## Remove correlated covariates

As noted above, removing three correlated covariates, plus BI since not going to
use now.

```{r, reduced}
input_river_reduced <- select(input_river,
                              -c("fraser_peak_spring",
                                 "total_ocean_salmon",
                                 # "total_spawners",
                                 "bi_anomaly"))
input_river_reduced %>% tail() %>% a()

corrplot::corrplot(cor(input_river_reduced,
                       use = "pairwise.complete.obs"),
  method = "number",
  type = "upper" # show only upper side
  )

```

## Age 4

Can take a while so saving the results as .rds
and not rerunning if the results are already saved. IF YOU WANT TO RERUN THE
MULTIVIEW EMBEDDING you need to set `rerun_age4 = TRUE` here.

```{r, red1}
rerun_age4 <- FALSE

# should rename to lags_reduced_age4 when time TODOo
lags_reduced_pdo <- list(age4 = c(1, 2, 3, 4),
                         effective_spawners = c(4),
                         pdo_winter_mean = c(2), # age4 just 2 year lag wrt return year as only 1.2
                         fraser_mean_spring = c(1, 2),  # not sure, could
                         # maybe even get 2026 value so have a lag of 0 in future
                         pink = c(2),  # not exactly sure, but dont' have 2025
                         # value yet so do 2 lag
                         sst_spring = c(2),  # same as PDO
                         sst_winter = c(2),
                         zoo_total_biomass = c(2))

if(rerun_age4){
  tictoc::tic()
  mve_reduced_pdo <- multiview_embedding(data = input_river_reduced,
                                         response = "age4",
                                         lags = lags_reduced_pdo)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()
  tictoc::toc()   # will be wrong if computer hibernated

  saveRDS(mve_reduced_pdo, file = "mve_reduced_pdo.rds")
} else {
  mve_reduced_pdo <- readRDS("mve_reduced_pdo.rds")
}
```

See the results
```{r, red1load}
mve_reduced_pdo$rho_prediction_from_mve
mve_reduced_pdo$rho_each_top_subset

plot_mve_spaces(mve_reduced_pdo,
                rho_on_top = FALSE)

# So there's lots, so need to just show a few of these:
# mve_reduced_pdo$lags_of_top_subsets[1:20]

# Look at the top few predictions:

# mve_reduced_pdo[["response_each_subset"]][mve_reduced_pdo$top_subsets_for_rho_index[1:5]]

mve_reduced_pdo$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r red1fig}
plot(input_river$age4,
     mve_reduced_pdo$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_reduced_pdo$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_reduced_pdo$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_reduced_pdo$response_predicted_from_mve[T+1], col = "red")
```




## Age 5

Remove lags that weren't important for age4 (see Chilko 2025 for more explanation),
else age5 will take about 6 hours to run.

2025 to reduce size, think we're okay for 2026: So looking at lags from the top 20 sets from age4 results above. Manually copy the code that defined them, then remove
ones that show up in top 20 sets, to leave ones not in top 20. Easiest to copy
`mve_reduced_pdo$lags_of_top_subsets[1:20]` into R to scroll through and search.

First set of age4 lags.
Not running:
```{r, lagscheck, eval = FALSE}
lags_reduced_pdo

lags_reduced_pdo_not_in_top_20 <- list(zoo_total_biomass = c(1, 2))

# Chilko had fraser_mean_spring lag 1 also
```

Can take a while so saving the results as .rds
and not rerunning if the results are already saved. IF YOU WANT TO RERUN THE
MULTIVIEW EMBEDDING you need to set `rerun_age5 = TRUE` here.

```{r, red3}
rerun_age5 <- FALSE
# PDO and sst: age5 2 year lag for 2.2 and 3 year lag (for 1.3)
lags_reduced_age5 <- list(age4 = c(1),
                          age5 = c(1, 2, 3, 4, 5),
                          effective_spawners = c(5),
                          pdo_winter_mean = c(2, 3),
                              fraser_mean_spring = c(1, 2, 3),
                              pink = c(2, 3),
                              sst_spring = c(2, 3))
                              # zoo_total_biomass = c(2, 3)) # not doing to
                              # reduce number of lags
                              # bi_anomaly = c(1, 2))
if(rerun_age5){
  tictoc::tic()
  mve_reduced_age5 <- multiview_embedding(data = input_river_reduced,
                                              response = "age5",
                                              lags = lags_reduced_age5)
  # Centre and scaling and first differencing get done automatically, in
  #  state_space_reconstruction_for_sve()
  tictoc::toc()
  
  saveRDS(mve_reduced_age5, file = "mve_reduced_age5.rds")
} else {
  mve_reduced_age5 <- readRDS("mve_reduced_age5.rds")
}
```

See results.

```{r, red3load}
mve_reduced_age5$rho_prediction_from_mve
mve_reduced_age5$rho_each_top_subset

# Taking too long (think as so many combinations to figure out), need to just
#get numbers TODO and in other rivers
#plot_mve_spaces(mve_reduced_age5,
#                rho_on_top = FALSE,
#                max_embeddings = 20)

# So there's lots, so need to just show a few of these:
# mve_reduced_age5$lags_of_top_subsets[1:20]

# Look at the top few predictions:

# mve_reduced_age5[["response_each_subset"]][mve_reduced_age5$top_subsets_for_rho_index[1:5]]

mve_reduced_age5$response_predicted_from_mve   # Overall mve prediction, including for T+1.
```

```{r red3fig}
plot(input_river$age5,
     mve_reduced_age5$response_predicted_from_mve[-(T+1)],
     xlab = "Age5 returns - observed",
     ylab = "Age5 returns - predicted")

legend("topleft", legend = c(paste("rho is",
                                    round(mve_reduced_age5$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_reduced_age5$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_reduced_age5$response_predicted_from_mve[T+1], col = "red")
```

## Summarise results

Summarise results of the two runs, and add them together (no need for picking
best like Chilko, as we haven't done a BI run here).

```{r, summarylim}
lim_common_max <- max(c(input_river$age4,
                        mve_reduced_pdo$response_predicted_from_mve,
                        input_river$age5,
                        mve_reduced_age5$response_predicted_from_mve),
                      na.rm = TRUE)
lim_common_max
lim_common = c(0, lim_common_max)
```

Panel plot, now adding in time series plots also. Left column plots have same
axes for age4 and age5, not so in right-hand time series.
```{r, summaryplot}
par(mfrow = c(2,2))

# Age4 correlation
plot(input_river$age4,
     mve_reduced_pdo$response_predicted_from_mve[-(T+1)],
     xlab = "Age4 returns - observed",
     ylab = "Age4 returns - predicted",
     main = "Age4",
     xlim = lim_common,
     ylim = lim_common)

legend("topleft", legend = c(paste("rho is",
                                   round(mve_reduced_pdo$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_reduced_pdo$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_reduced_pdo$response_predicted_from_mve[T+1], col = "red")

# Age4 time series
plot(input_river$year,
     input_river$age4,
     type = "o",
     main = "Age4",
     xlab = "Year",
     ylab = "Age4 returns")
     # ylim = lim_common)

points(c(input_river$year, max(input_river$year) + 1),
       mve_reduced_pdo$response_predicted_from_mve,
       pch = 16,
       col = "red",
       cex = 0.8)

legend("topleft",
       legend = c("observed", "predicted"),
       pch = c(1, 16),
       col = c("black", "red"))

# Age5 correlation
plot(input_river$age5,
     mve_reduced_age5$response_predicted_from_mve[-(T+1)],
     xlab = "Age5 returns - observed",
     ylab = "Age5 returns - predicted",
     main = "PDO age5",
     xlim = lim_common,
     ylim = lim_common)

legend("topleft", legend = c(paste("rho is",
                                   round(mve_reduced_age5$rho_prediction_from_mve, 2)),
                              paste("prediction is",
                                    f(mve_reduced_age5$response_predicted_from_mve[T+1]))))
abline(a = 0, b = 1)
abline(h = mve_reduced_age5$response_predicted_from_mve[T+1], col = "red")

# Age5 time series
plot(input_river$year,
     input_river$age5,
     type = "o",
     main = "Age5",
     xlab = "Year",
     ylab = "Age5 returns")
     # ylim = lim_common)

points(c(input_river$year, max(input_river$year) + 1),
       mve_reduced_age5$response_predicted_from_mve,
       pch = 16,
       col = "red",
       cex = 0.8)

legend("topleft",
       legend = c("observed", "predicted"),
       pch = c(1, 16),
       col = c("black", "red"))
```

Summarise, and then sum together age4 and age5 to get our prediction.

```{r, summarise}
summary <- tibble(age = "age4",
                  run = "PDO",
                  rho = mve_reduced_pdo$rho_prediction_from_mve ,
                  prediction = mve_reduced_pdo$response_predicted_from_mve[T+1])
summary <- rbind(summary,
                 c("age5",
                   "PDO",
                   mve_reduced_age5$rho_prediction_from_mve,
                   mve_reduced_age5$response_predicted_from_mve[T+1])) %>%
  dplyr::mutate_at(vars(age,
                        run),
                   factor) %>%
  dplyr::mutate_at(vars(rho,
                        prediction),
                   as.numeric)

summary

knitr::kable(summary,
             digits = c(NA, NA, 2, 6))
stopifnot(nrow(summary) == 2)

total_returns_prediction <- sum(summary$prediction)
```

The prediction for combined age4 and age5 returns `r river` is ... `r f(total_returns_prediction)`.


